---
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# noaaquakes

The noaaquakes package provides a suite of tools for reading, cleaning, and visualizing the NOAA Significant Earthquake Database. It includes specialized ggplot2 geometries for timelines and leaflet wrappers for interactive mapping.

<!-- badges: start -->
[![R-CMD-check](https://github.com/derneld/noaaquakes/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/derneld/noaaquakes/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->


## Installation

You can install the development version of noaaquakes from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("derneld/noaaquakes")
```

## Data Preparation

The package provides functions to import the raw NOAA TSV file and convert it into a tidy format suitable for analysis.

```{r example1}
library(noaaquakes)
library(dplyr)

# Find the correct path regardless of whether the package is installed or in development
file_path <- system.file("extdata", "earthquakes.tsv", package = "noaaquakes")

# If system.file returns empty (package not installed yet), point to the local file
if (file_path == "") {
  file_path <- "inst/extdata/earthquakes.tsv"
}

raw_data <- eq_read_data(file_path)
clean_data <- eq_clean_data(raw_data)
```

## Visualizing Regional Earthquake Activity

The package introduces geom_timeline() to plot earthquakes on a horizontal time axis, with the ability to stratify by region (y-axis) and scale by magnitude or deaths.

```{r example2}
library(ggplot2)

clean_data %>%
  filter(COUNTRY %in% c("USA", "China", "Mexico"), Year > 2000) %>%
  ggplot(aes(x = DATE, y = COUNTRY, color = DEATHS, size = EQ_PRIMARY)) +
  geom_timeline() +
  theme_minimal()
```

### Adding Labels

You can use geom_timeline_label() to highlight the most significant events by magnitude.

```{r example3}
clean_data %>%
  filter(COUNTRY == "USA", Year > 2010) %>%
  ggplot(aes(x = DATE, y = COUNTRY, color = DEATHS, size = EQ_PRIMARY)) +
  geom_timeline() +
  geom_timeline_label(aes(label = LOCATION_NAME, magnitude = EQ_PRIMARY), n_max = 3) +
  theme_minimal()
```

## Interactive Maps

The eq_map() function leverages leaflet to create interactive visualizations. You can generate custom HTML labels using eq_create_label() to show detailed information in pop-ups.

```{r example4, screenshot.force = TRUE}
clean_data %>%
  filter(COUNTRY == "Mexico" & Year > 2010) %>%
  mutate(popup_text = eq_create_label(.)) %>%
  eq_map(annot_col = "popup_text")
```

